Symmetric Peer-to-Peer Applications

### 2022-10-18 tpd-22.tex
...

\IEEEtitleabstractindextext{%
\begin{abstract}
In collaborative networked applications, such as shared online documents,
users can interact remotely and yet share the exact same experience as if they
were in a single machine.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a restricted API, which is deterministic,
stateless, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, ased on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their state.
%
TODO: application, results
\end{abstract}

@@

\IEEEPARstart{C}{ollaborative} networked applications allow multiple users to
interact remotely and yet share the same experience in real time.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
shared documents, watch parties, and multi-player games.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline.
However, current solutions depend on a central server to interconnect network
nodes and determine the shared timeline.

In this work, we propose \emph{symmetric peer-to-peer applications}, which
do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer networks.
Events are flooded in the graph and are triggered locally with a small delay
to compensate the network latency.
To deal with events received out of order or too late, a distributed time
machine can rollback peers to a previous state and reapply events in order and
in time.
Our main contribution is the design of a middleware that runs on each peer to
ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

TODO: application, results

@@

In this section, we revisit existing solutions for symmetric distributed
applications.
We focus on
    (a) how the network is organized,
    (b) how global time is determined, and
    (c) how events are propagated and applied.
Figure~\ref{fig.table} compares selected works regarding these aspects.

Croquet%
\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
bit-identical real-time behavior for every user in a collaborative
distributed environment.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
periodic ticks, such that all nodes in the network advance together according
to a synchronized global shared clock.
If a user generates an event, it is delayed and sent back to the reflector,
which broadcasts the event in the next tick, which all nodes apply in sync.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to restore the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network, in which nodes advance in sync, and in which event outcomes depend
on the central server to be online.

GALS~\cite{gals} shares the same goals with Croquet, but with some tradeoffs,
mostly favoring clients with slow connections.
As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
the server, clients have their own independent local clocks.
Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
delayed dynamically according to the slowest client.
%
On the one hand, ticks do not generate any traffic and clients have smooth
frame transitions, even those with poor connections.
On the other hand, events are slower and clients may experience occasional
freezes.
The syncing protocol also requires bookkeeping to deal with clock drifts and
client disconnections~\cite{gals}.
%
As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
network, but in which nodes advance time independently, and in which event
outcomes still depend on the central server to be online.

In both solutions, the network advances together as a whole in real time, with
a total order among events, which is determined by a central server that must
be permanently online.
All clients compute the events in sequence, respecting their timestamps, and
using only deterministic and stateless calls.
This way, it is guaranteed that all clients reach the same state, and
observe bit-identical steps.

An antagonistic approach to deterministic reactions to events is to model the
application with conflict-free replicated data types (CRDTs)~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the peers are
offline.
%
Since operations must be commutative, it is not possible to timestamp them
in a unique global shared clock.
Hence, there is no notion of a timeline in which peers go through
bit-identical steps, but they do eventually reach the same state.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks,
in which peers advance independently, and in which operations can be applied
immediately, even while offline.
%
Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
is an example of an implementation of decentralized
CRDTs~\cite{p2p.automerge,p2p.pushpin}.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
peer-to-peer real-time symmetric execution, while still tolerating offline and
out-of-order event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Events are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
Like Croquet and GALS, peers compute events in sequence using only
deterministic and stateless calls.

Regarding decentralized time machines, Fusion%
\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
is a state synchronization networking library for the Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.

Regarding single-user time machines, the video game Braid%
\footnote{\url{Braid: http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating mutation from
rendering~\cite{tml.alive}.

@@

In this section, we describe the architecture of our middleware and also the
API in which developers have to conform with.
%
The middleware is written in C and relies on the library \emph{SDL}%
\footnote{SDL: \url{www.libsdl.org}}
to deal with the network, timers, and input events, such as mouse clicks and
key presses.
The full source code%
\footnote{TODO}
is under 500 lines of code.

@@

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users use the arrow keys to control the direction of a
moving rectangle.
The six instances are connected in a peer-to-peer network, such that some
peers can only communicate indirectly.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame with the rectangle being ``bit-identically'' at
the same position.

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a delay.
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
the other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp scheduled to the future, such that all
peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate delay so that they can reach other peers in time to be applied in
sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer that is locally ahead of time may receive an event that
should have been applied in the past, even considering the delay.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
\item \textbf{Peer Churn:}
Considering that we target peer-to-peer networks, it is important that the
middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that the
peer receives all events that ever happened in order, which is trivial since
they all hold an absolute timestamp in the shared timeline.
\end{enumerate}

@@

The application generates events that need to be broadcast to the network so
that all peers behave the same.
Note that the application events may (or may not) differ from low-level local
system events.
%which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.
%
The middleware predefines two events for all applications:
    \code{P2P\_EVT\_START} represents the beginning of the simulation, and
    \code{P2P\_EVT\_TICK} is generated every frame.
The application-specific events must be declared by the programmer in an
enumeration.
In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
presses:

{\footnotesize
\begin{verbatim}
enum {
    P2P_EVT_KEY = P2P_EVT_NEXT
};
\end{verbatim}
}

@@

The middleware calls \code{cb\_ini} twice: at the beginning and at the end of
the loop.
The callback initializes (and finalizes) the network topology, as well as
static immutable globals that can live outside the simulation memory, such as
the window and image textures in our example:

@@

The middleware calls \code{cb\_sim} every frame, which affects the simulation
state \code{G} deterministically.
The occurring event is passed as an argument to the callback, which must never
perform any side effects, such as stateful calls or rendering frames.
In our example, we modify the state of the rectangle as follows:
    (a) reset its position and speed on \code{START},
    (b) increment its position based on the current speed on \code{TICK}, and
    (c) modify its speed based on \code{KEY}:

@@

The callback \code{cb\_evt} is called by the middleware in real time, whenever
a local SDL event occurs.
This callback has two goals:
    (a) map and broadcast application events, and
    (b) provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to broadcast application events in the
network.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 key events, and
we also exhibit the pressed key in real time as a visual feedback:

@@

{\footnotesize
\begin{verbatim}
01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
02    cb_ini(1);    // initialization
03    cb_sim(<P2P_EVT_START>);
04
05    while (<app-running>) {
06      if (<next-network-packet>) {
07        // fast forward if I'm late
08        if (<packet-sync-future>) {
09          p2p_travel(<now>, <packet-sync-time>, <speed>);
10        }
11
12        // travel back & forth if I'm early
13        if (<packet-evt-past>) {
14          p2p_travel(<now>, <packet-evt-time>, <speed>);
15          p2p_travel(<packet-evt-time>, <now>, <speed>);
16        }
17
18        // real time if I'm on time
19        if (<packet-evt-ok>) {
20          cb_sim(<packet-evt>);
21          cb_ren();
22        }
23      }
24
25      if (<next-local-input>) {
26        // broadcast local event
27        p2p_evt evt;
28        if (cb_evt(&evt)) {
29          p2p_bcast(<future>, &evt);
30        }
31      }
32
33      // simulate tick in real time
34      if (<next-local-tick>) {
35        cb_sim(<P2P_EVT_TICK>);
36        <application-snapshot-every-second>
37        cb_ren();
38      }
39    }
40
41    cb_ini(0);    // finalization
42  }
\end{verbatim}
}

@@

In this section, we evaluate if (and to what extent) our design meets its goals
to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
time, and (c) can leave and join the network and remain symmetric.
%
We simulate an application varying the network latency and peer churn, as well
as the frequency and deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.

@@

We simulate the application with $21$ peers using the mixed topology of
Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
The topology has an average of 5 hops between peers (e.g., \code{4--16} are 7
hops away).
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms
distribution normal}).
%
We vary
    (i)   the network latency (\code{5--500ms} between each peer),
    (ii)  the rate of events (\code{5--200 evt/sec} in the whole network), and
    (iii) the delay multiplier (\code{1--5x} considering the latency and the
          average hops between peers).
Regarding item (c), a \code{50ms} latency with a \code{2x} multiplier results
in \code{500ms} event delays, i.e., each event is triggered with this delay to
compensate the given latency and fixed average of 5 hops.
%
These variations result in $108$ combinations.
We executed each combination $3$ times for $5$ minutes each.
Therefore, the experiments take $27$ hours to complete.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{baks}
  \caption{
    \label{fig.baks}
The \code{5x}, \code{2x}, and \code{1x} delay regions each contains $36$
measures for multiple configurations of event rate (\code{5--200evt/min} and
network latency (\code{5--500ms}).
%
The color thresholds (\emph{white}, \emph{yellow}, \emph{red}, \emph{black})
are arbitrary but help to distinguish between regions of interest.
%
The experiments uses $1$ process for each peer in a single Linux machine
(\emph{i7/16GB}).
\end{figure*}

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
In real-time networked applications, such as shared documents, users can
interact remotely and yet share the exact same experience as if they were in a
single machine.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a restricted API, which is deterministic,
stateless, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, based on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their state.
%
We show that the middleware can handle applications with over 20 peers in a
heterogeneous topology under high churn.
For instance, in a simulation scenario of 25 events per minute with a network
latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
the time.
\end{abstract}

@@

\IEEEPARstart{R}{eal-time} networked applications allow multiple users to
interact remotely and yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
shared documents, watch parties, and multi-player games.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline.
However, current solutions depend on a central server to interconnect network
nodes and determine a shared timeline.

In this work, we propose \emph{symmetric peer-to-peer applications}, which
do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer networks.
Events are flooded in the graph and are triggered locally with a small delay
to compensate the network latency.
To deal with events received out of order or too late, a distributed time
machine can rollback peers to a previous state and reapply events in order and
in time.
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
and straight lines.
We vary the network latency and peer churn, as well as the frequency and
deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.
%
As results, we show that the last two goals are met even under extreme
conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
$1.5s$ to recover from churns.
%
For the first goal, we identify the scenarios in which the middleware can meet
event deadlines with the desired performance.
As an example, in a scenario of 25 events per minute with a network latency of
$50ms$, the peers need to roll back only $3\%$ of the time.

@@

In this section, we revisit existing solutions for symmetric distributed
applications.
We focus on
    (a) how the network is organized,
    (b) how global time is determined, and
    (c) how events are propagated and applied.
Figure~\ref{fig.table} compares three selected works regarding these aspects.

Croquet%
\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
bit-identical real-time behavior for users in collaborative distributed
environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
periodic ticks, such that all nodes in the network advance together according
to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to restore the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event outcomes
depend on the central server to be online (c).

GALS~\cite{gals} shares the same goals with Croquet, but with some tradeoffs,
mostly favoring clients with slow connections.
As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
the server, clients have their own independent local clocks.
Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
delayed dynamically according to the slowest client.
%
On the one hand, ticks do not generate any traffic and clients have smooth
frame transitions, even those with poor connections.
On the other hand, events take longer to apply and clients may experience
occasional freezes.
The syncing protocol also requires extra bookkeeping to deal with clock drifts
and client disconnections~\cite{gals}.
%
As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
network (a), but in which nodes advance time independently (b), and in which
event outcomes still depend on the central server to be online (c).

In both solutions, the network advances together as a whole in real time, with
a total order among events, which is determined by a central server that must
be permanently online.
All clients compute events in sequence, respecting timestamps, and using only
deterministic and stateless calls.
This way, it is guaranteed that all clients reach the same state, and
observe bit-identical steps.

An antagonistic approach to deterministic reactions to events is to model the
application with conflict-free replicated data types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the peers are
offline.
%
Since operations must be commutative, it is not possible to timestamp them
in a unique global shared clock.
Hence, there is no notion of a timeline in which peers go through
bit-identical steps, but they do eventually reach the same state.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
(a), in which peers advance independently (b), and in which operations can be
applied immediately, even while offline (c).
%
Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
peer-to-peer real-time symmetric execution (a), while still tolerating
out-of-order (b) and offline (c) event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Events are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
Like Croquet and GALS, peers compute events in sequence using deterministic and
stateless calls only.

We now discuss related work on time machines.
%
Regarding decentralized machines, Fusion%
\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
is a state synchronization networking library for the Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.
%
Regarding single-user local machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state modification
from rendering~\cite{tml.alive}.

@@

The middleware is written in C and relies on the library \emph{SDL}%
\footnote{SDL: \url{http://www.libsdl.org}}
to deal with the network, timers, and input events, such as mouse clicks and
key presses.
The full source code%
\footnote{TODO}
is under 500 lines of code.

@@

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users use the arrow keys to control the direction of a
moving rectangle.
The six instances are connected in a peer-to-peer network, such that most
peers can only communicate indirectly to each other.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame with the rectangle being ``bit-identically'' at
the same position.

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a $\Delta$ delay.
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
an API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a $\Delta$ delay scheduled to the
future, such that all peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate $\Delta$ so that they can reach other peers in time to be applied in
sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, even considering the delay.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
\item \textbf{Peer Churn:}
Considering that we target peer-to-peer networks, it is important that the
middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)), including their absolute timestamps in the shared
timeline.
\end{enumerate}

@@

When an application generates events, they need to be broadcast to the network
so that all peers behave the same.
Note that application events may (or may not) differ from low-level local
system events.
%which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.
%
The middleware predefines two events for all applications:
    \code{P2P\_EVT\_START} represents the beginning of the simulation, and
    \code{P2P\_EVT\_TICK} is generated every frame.
The other application-specific events must be declared by the programmer in an
enumeration.
In our example, we define a new event \code{P2P\_EVT\_KEY} to represent key
presses:

{\footnotesize
\begin{verbatim}
enum {
    P2P_EVT_KEY = P2P_EVT_NEXT // next to predefined
};
\end{verbatim}
}

@@

The middleware calls \code{cb\_ini} twice: at the beginning and at the end of
the loop.
The callback should initialize (and finalize) the network topology, as well as
static immutable globals that can live outside the simulation memory, such as
the window and image textures in our example:

@@

The middleware calls \code{cb\_sim} every frame or event occurrence, which
should affect the simulation state \code{G} deterministically.
An occurring event may be passed as argument to the callback, which must never
perform any side effects, such as stateful calls or rendering frames.
In our example, we modify the state of the rectangle as follows:
    (a) reset its position and speed on \code{START},
    (b) increment its position based on the current speed on \code{TICK}, and
    (c) modify its speed based on \code{KEY}:

@@

The callback \code{cb\_evt} is called by the middleware in real time, whenever
a local SDL event occurs.
This callback has two goals:
    (a) map and broadcast application events, and
    (b) provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to broadcast application events in the
network.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 key events of
interest, but we also exhibit the pressed key in real time as a visual
feedback:

@@

{\footnotesize
\begin{verbatim}
01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
02    cb_ini(1);    // initialization
03    cb_sim(<P2P_EVT_START>);
04
05    while (<app-running>) {
06      if (<next-network-packet>) {
07        // fast forward if I'm late
08        if (<packet-sync-future>) {
09          p2p_travel(<now>, <pak-sync-time>, <vel>);
10        }
11
12        // travel back & forth if I'm early
13        if (<packet-evt-past>) {
14          p2p_travel(<now>, <pak-evt-time>, <vel>);
15          p2p_travel(<pak-evt-time>, <now>, <vel>);
16        }
17
18        // real time if I'm on time
19        if (<packet-evt-ok>) {
20          cb_sim(<packet-evt>);
21          cb_ren();
22        }
23      }
24
25      if (<next-local-input>) {
26        // broadcast local event
27        p2p_evt evt;
28        if (cb_evt(&evt)) {
29          p2p_bcast(<future>, &evt);
30        }
31      }
32
33      // simulate tick in real time
34      if (<next-local-tick>) {
35        cb_sim(<P2P_EVT_TICK>);
36        <application-snapshot-every-second>
37        cb_ren();
38      }
39    }
40
41    cb_ini(0);    // finalization
42  }
\end{verbatim}
}

@@

In this section, we evaluate if (and to what extent) our design meets its goals
to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
time, and (c) can leave and join the network and remain symmetric.
%
We perform simulation varying the network latency and peer churn, as well as
the frequency and deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=2in]{mult}
  \caption{
    \label{fig.mult}
Event delays in $ms$ considering the network latency, delay multiplier, and the
average of 5 hops between peers from Figure~\ref{fig.topo}.
    }
\end{figure}

We perform simulations with 21 peers using the mixed topology of
Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
The topology has an average of 5 hops between peers (e.g., \code{4--16} are 7
hops away).
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution.
% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
%
We vary
    (i)   the network latency (\code{5--500ms} between each peer),
    (ii)  the rate of events (\code{5--200 evt/sec} in the whole network), and
    (iii) the delay multiplier (\code{1--5x} considering the latency and the
          average hops between peers).
%
Figure~\ref{fig.mult} shows the event delays considering the network latency
and chosen multiplier.
For instance, a \code{50ms} latency with a \code{2x} multiplier results in
\code{500ms} event delays, i.e., each event is triggered with this delay to
compensate the given latency and fixed average of 5 hops.
%
These variations result in $108$ combinations.
We executed each combination $3$ times for $5$ minutes each.
We also make some modifications to evaluate peer churn, which requires to
re-execute the simulation.
The experiments take around $40$ hours to complete.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{baks}
  \caption{
    \label{fig.baks}
The \code{5x}, \code{2x}, and \code{1x} delay regions each contains $36$
measures for multiple configurations of event rate (\code{5--200evt/min} and
network latency (\code{5--500ms}).
%
The color thresholds (\emph{white}, \emph{yellow}, \emph{red}, \emph{black})
are arbitrary but help to distinguish between regions of interest.
%
The experiments uses $1$ process for each peer in a single Linux machine
(\emph{i7/16GB}).
    }
\end{figure*}

We now detail the results and how we instrument the middleware implementation
to measure the properties above:
%
\begin{enumerate}
%
\item \textbf{Recurrence of time travels:}
Every time a peer is ahead of time and needs to travel back, we output its id
and how much ticks it needs to travel back.
At the end, we measure the percentage of the sum of all time travel ticks over
the total simulation ticks.
As an example, if all peers sum $1000$ time travel ticks over $100k$
accumulated simulation ticks, then the final measure is $1\%$.
%
Figure~\ref{fig.baks} presents the results considering all combinations of
parameters (i), (ii), and (iii) above.
The colors indicate the feasibility of each tested scenario
    (\emph{white=good}, \emph{yellow=moderate}, \emph{red=poor},
    \emph{black=faulty}).
%
For instance, the three framed rectangles ($0.1$, $3.1$, and $7.2$) represent
the scenarios with $50ms$ network delay, generating $25evt/min$.
Each rectangle uses a different delay multiplier, showing a good performance
for $5x$, moderate for $2x$, and poor for $1x$.
%
\item \textbf{Real-time pace of peers:}
All peers output their current ticks continuously, allowing us to compare them
in the shared timeline.
Every time we see a greater tick never seen in the timeline before, we count
smaller ticks from other peers, which constitute time violations.
As an example, if all peers sum $1000$ of such smaller ticks over $100k$
accumulated simulation ticks, then the final measure is $1\%$.
%
We also start each peer with a random delay of up to $10s$ to force a
synchronization mismatch at the beginning.
%
Considering all scenarios, the average is negligibly under $0.2\%$, with no
significant variance.
Therefore, we omit the resulting table akin to Figure~\ref{fig.baks}), which
would not provide further insights.
%
\item \textbf{Correctness of unstable peers:}
We tweak our simulation to make each peer remain $40s$ online and $20s$
offline in the average, including peers \code{9--11} in the middle of the
network.
This creates partitions in the network, which generate events out of sync that
require constant resynchronization between peers.
%
Then, to evaluate the network correctness under such high churn, we count how
much time the recovering peers take to catch up in real time:
When a peer becomes online, we take the current maximum tick considering all
reachable peers, and then count the time the recovering peer takes to catch up.
We also assert that they all reach the same final state eventually.
%
Considering all scenarios, the average is $1.3$ seconds of recovering time,
with no significant variance.
Therefore, we also omit the resulting table akin to Figure~\ref{fig.baks}).
%
\end{enumerate}

\begin{comment}
$ vi test.c: // #define TEST_OFFLINE
$ sudo ./test-ana.sh 2>&1 | tee all.log
$ ./test-sheet.sh > sheet.log
$ ./test-sheet-2.sh > sheet-2.log
$ lua5.3 test-sheet-2.lua sheet-2.log > /tmp/x
$ gedit /tmp/x

$ vi test.c: #define TEST_OFFLINE
$ sudo ./test2-ana.sh 2>&1 | tee all2.log
$ lua5.3 test2-sheet.lua
\end{comment}
